Gringo 2.0

### 2008-09-03 doc/documentation/paper.tex
Doku: Added intuitive description of lamdba-restrictedness, removed examples, restructured whole document
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@485 3e02f596-3131-0410-8352-e04eeaec37cc

\section{$\lambda$-Restriction}
\label{domainrestricted}
There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs. For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
gringo2.0 verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs. The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.

@@

@@

 :- not on(X+1,X,k), block(X), X < n.
\end{verbatim}
To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
A potential command line call could look like
\begin{verbatim}
 iclingo --const n=10 blocksworld.lp
\end{verbatim}
and the answer to the problem can be read from the answer set.
To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
\begin{verbatim}
 #hide.
 #show move/3.
\end{verbatim}

\paragraph{const}
Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
while they are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
It has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
The compute statement is solver dependent and not supported by every solver.
\paragraph{comments}
Comments included in the input are line based.
Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
\subsubsection{Meta-Syntax for Incremental Programs}
Incremental programs must be splitted into 3 parts.
The base part, for the information that hold in all steps ans is independent of the incremental constant.
The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
And finally the volatile part. In this part the queries to the program are given, that must hold in the last state.
To seperate the parts, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
To start the base part
\begin{verbatim}
 #base.
\end{verbatim}
must be written. Now all rules of the base part can follow.
The use of the incremental constant is not allowed in this part.
For the cumulative part an additional constant has to be passed to the statement.
\begin{verbatim}
 #cumulative k.
\end{verbatim}
Now all incremental rules follow. The incremental constant used for grounding is $k$.
The same holds for the volatile part.
\begin{verbatim}
 #volatile k.
\end{verbatim}
The ``\#''-sign can not be omitted for these meta statements.
These statements can also be mixed, only the last statement is important.
If no statement is given at the beginning of the file, base is assumed.

@@

\section{Examples}
In this section some encodings for common problems are presented and explained.
\paragraph{Fastfood}
This example needs a proper instance to work.
\begin{verbatim}
#hide.
#show altdepot/2.
#show costs/1.

% calculate the number of depots
depots(X) :- X = { depot(DN, DK) :  depot(DN, DK) }.

% calculate the costs for the initial depots
% (the grounder can calculate this on its own)
costs(C) :- C = [ serves(RN, Dist) : restaurant(RN, RK) : dist(Dist) = Dist ].

% choose the closest depot for a restaurant
% (the grounder can calculate this on its own)
serves(Rname,Dist) :- restaurant(Rname,RK),
   Dist = min [ depot(DN1, DK1) : depot(DN1, DK1) = abs(DK1-RK) ].

dist(abs(X-Y)) :- restaurant(A, X), restaurant(B, Y).

% Each restaurant may be an alternative depot or not.
X { altdepot(R,K) : restaurant(R,K) } X :- depots(X).

% choose the closest depot for a restaurant
altserves(Rname,Dist) :- restaurant(Rname,RK),
   Dist = min [ altdepot(DN1, DK1) : restaurant(DN1, DK1) = abs(DK1-RK) ].

% fail if we have not found a better solution
:- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].

% thats the easiest way to show the calculated costs
minimize [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
\end{verbatim}

\paragraph{Queens}
\begin{verbatim}
#hide.
#show q(X,Y).

% this can be set via the command line
#const queens = 200.

#const diags  = queens * 2 - 1.
d(1..queens).
diag(1..diags).

% enumerate the diagonals
diagX(X,Y,X - Y + queens) :- d(X), d(Y).
diagY(X,Y,X + Y - 1) :- d(X), d(Y).

% exactly 1 queen in each row/col
1 { q(X,Y) : d(Y) } 1 :- d(X).
1 { q(X,Y) : d(X) } 1 :- d(Y).

% at most 1 queen on each diagonal
{ q(X,Y) : diagX(X,Y,D) } 1 :- diag(D).
{ q(X,Y) : diagY(X,Y,D) } 1 :- diag(D).

>>>>

@@

\subsection{$\lambda$-Restriction}
\label{domainrestricted}
There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs. For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
gringo2.0 verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs. The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
\newline
To give a gross impression of what $\lambda$-restriction means for writing a logic program an informal description is given.
In gringo2.0 every variable used in a rule must have a complete domain. This means that this variable has to occur in at least one predicate whichs domain is completed. Therefore the predicate must be in the head of rules that are already domain restricted. So the grounder has to find a valid order of the rules so that every predicate in the head is completed before it is used as a domain predicate for the next rules.
To make things more clear, an example is given
\begin{verbatim}
 zig(0) :- not zag(0).   zag(0) :- not zig(0).
 zig(1) :- not zag(1).   zag(1) :- not zig(1).
 zigzag(X,Y) :- zig(X), zag(Y).
\end{verbatim}
The rules of this example can be ordered in the following way.
At first, the rules with \texttt{zig/1} and \texttt{zag/1} are considered.
Since these are the only rules \texttt{zig/1} and \texttt{zag/1} occurs in the head, the domains of the predicates are closed. The domains \{\texttt{zig(0)}, \texttt{zig(1)}\} and \{ \texttt{zag(0)}, \texttt{zag(1)}\} are complete, since no other predicates of the form \texttt{zig/1} or \texttt{zag/1} can be derived from the program.
Now the third rule with \texttt{zigzag/2} in the head can be grounded. The variables \texttt{X} and \texttt{Y} both occur in a domain predicate, and the values $0$ respectively $1$ can substitute the variables.
\newline
To give a comparison to the $\omega$-restricted programs of lparse, an extra domain predicate has to be introduced to the example.
\begin{verbatim}
 domain(0..1).
 zig(0) :- not zag(0).   zag(0) :- not zig(0).
 zig(1) :- not zag(1).   zag(1) :- not zig(1).
 zigzag(X,Y) :- zig(X), zag(Y), domain(X), domain(Y).
\end{verbatim}
\texttt{zig/1} and \texttt{zag/1} can not be used as a domain predicate, since their truth value can not be fully evaluated.

@@

@@

\paragraph{const}
Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
while they are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
It has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
The compute statement is solver dependent and not supported by every solver.
\paragraph{comments}
Comments included in the input are line based.
Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
TODO: Block comments
\subsection{Meta-Syntax for Incremental Programs}
Incremental programs must be splitted into 3 parts.
The base part, for the information that hold in all steps ans is independent of the incremental constant.
The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
And finally the volatile part. In this part the queries to the program are given, that must hold in the last state.
To seperate the parts, three statement exists. Everything beyond such a statement is considered to belong to the specific part.
To start the base part
\begin{verbatim}
 #base.
\end{verbatim}
must be written. Now all rules of the base part can follow.
The use of the incremental constant is not allowed in this part.
For the cumulative part an additional constant has to be passed to the statement.
\begin{verbatim}
 #cumulative k.
\end{verbatim}
Now all incremental rules follow. The incremental constant used for grounding is $k$.
The same holds for the volatile part.
\begin{verbatim}
 #volatile k.
\end{verbatim}
The ``\#''-sign can not be omitted for these meta statements.
These statements can also be mixed, only the last statement is important.
If no statement is given at the beginning of the file, base is assumed.

\section{Examples}
In this section some encodings for common problems are presented and explained.

 :- not on(X+1,X,k), block(X), X < n.
\end{verbatim}
To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
A potential command line call could look like
\begin{verbatim}
 iclingo --const n=10 blocksworld.lp
\end{verbatim}
and the answer to the problem can be read from the answer set.
To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
\begin{verbatim}
 #hide.
 #show move/3.
\end{verbatim}
%\paragraph{Fastfood}
%This example needs a proper instance to work.
%\begin{verbatim}
%#hide.
%#show altdepot/2.
%#show costs/1.
%
%% calculate the number of depots
%depots(X) :- X = { depot(DN, DK) :  depot(DN, DK) }.
%
%% calculate the costs for the initial depots
%% (the grounder can calculate this on its own)
%costs(C) :- C = [ serves(RN, Dist) : restaurant(RN, RK) : dist(Dist) = Dist ].
%
%% choose the closest depot for a restaurant
%% (the grounder can calculate this on its own)
%serves(Rname,Dist) :- restaurant(Rname,RK),
%   Dist = min [ depot(DN1, DK1) : depot(DN1, DK1) = abs(DK1-RK) ].
%
%dist(abs(X-Y)) :- restaurant(A, X), restaurant(B, Y).
%
%% Each restaurant may be an alternative depot or not.
%X { altdepot(R,K) : restaurant(R,K) } X :- depots(X).
%
%% choose the closest depot for a restaurant
%altserves(Rname,Dist) :- restaurant(Rname,RK),
%   Dist = min [ altdepot(DN1, DK1) : restaurant(DN1, DK1) = abs(DK1-RK) ].
%
%% fail if we have not found a better solution
%:- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
%
%% thats the easiest way to show the calculated costs
%minimize [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].
%\end{verbatim}
%
%\paragraph{Queens}
%\begin{verbatim}
%#hide.
%#show q(X,Y).
%
%% this can be set via the command line
%#const queens = 200.
%
%#const diags  = queens * 2 - 1.
%d(1..queens).
%diag(1..diags).
%
%% enumerate the diagonals
%diagX(X,Y,X - Y + queens) :- d(X), d(Y).
%diagY(X,Y,X + Y - 1) :- d(X), d(Y).
%
%% exactly 1 queen in each row/col
%1 { q(X,Y) : d(Y) } 1 :- d(X).
%1 { q(X,Y) : d(X) } 1 :- d(Y).
%
%% at most 1 queen on each diagonal
%{ q(X,Y) : diagX(X,Y,D) } 1 :- diag(D).
%{ q(X,Y) : diagY(X,Y,D) } 1 :- diag(D).
%
%\end{verbatim}
